---
title: "078_Upload"
author: "Matteo Gasparotto"
date: "2025/05/08"
description: ""
output: 
  bookdown::html_document2:
    code_folding: hide
    fig_caption: true
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: yes
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy.opts=list(width.cutoff=80),tidy=TRUE, fig.asp=0.5, fig.width=12, warning = FALSE, message=FALSE)
```

```{r loadLibraries, message=FALSE, warning=FALSE}
source("~/R/Set_Environment.R")
source("~/R/Set_functions.R")
options(stringsAsFactors = FALSE)
mem.maxVSize(16384*6) 
set.seed(42)
options(future.globals.maxSize = 3 * 1024^3)
```

# Integration

```{r}
df1<-readRDS("../data/CTRL1.rds")
df2<-readRDS("../data/CTRL2.rds")
df3<-readRDS("../data/065_mild_2.rds")
df4<-readRDS("../data/065_mod_2.rds")
df5<-readRDS("../data/065_sev_1.rds")
df6<-readRDS("../data/065_sev_2.rds")
```

```{r}
obj <- c("df1", "df2", "df3", "df4", "df5", "df6")
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

for (o in obj) {
  tmp <- get(o)
  DefaultAssay(tmp)<-"RNA"
  tmp <- CellCycleScoring(tmp, s.features = s.genes, g2m.features = g2m.genes)
  tmp <- SCTransform(tmp, vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"))
  assign(o, tmp)
}
```


```{r}
dfs<-list(DS1=df1, DS2=df2, DS3=df3, DS4=df4, DS5=df5, DS6=df6)
features <- SelectIntegrationFeatures(object.list = dfs, nfeatures = 3000)
dfs <- PrepSCTIntegration(object.list = dfs, anchor.features = features)

anchors <- FindIntegrationAnchors(
  object.list = dfs,
  normalization.method = "SCT",
  anchor.features = features,
  dims = 1:30
)

df <- IntegrateData(
  anchorset = anchors,
  normalization.method = "SCT",
  dims = 1:30
)
rm(ctrl, mild, mod, sev, tmp, df1,df2,df3,df4,df5,df6,df7, dfs)
gc()
```

# Dimensionality reduction

## Elbowplot

```{r}
DefaultAssay(df) <- "integrated"
df <- CellCycleScoring(df, s.features = s.genes, g2m.features = g2m.genes)
df <- ScaleData(df, verbose = TRUE, vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"))
df <- RunPCA(df, npcs = 50, verbose = FALSE)
p1<-ElbowPlot(df, ndims=50)
p1 <- p1 + geom_vline(xintercept = 30, linetype = "dashed", color = "#025e8d", size = 1)
p1
```
```{r}
stdev <- df[["pca"]]@stdev
variance_frac   <- stdev^2 / sum(stdev^2)
cumulative_frac <- cumsum(variance_frac)

elbow_df <- data.frame(
  PC              = seq_along(variance_frac),
  variance_frac   = variance_frac,
  cumulative_frac = cumulative_frac
)[1:50, ]

d1       <- elbow_df$variance_frac
d2       <- diff(d1)
d3       <- diff(d2)
elbow_pc <- which.max(-d3) + 1  

p <- ggplot(elbow_df, aes(x = PC)) +
  geom_bar(aes(y = variance_frac), stat = "identity", alpha = 0.6) +
  geom_line(aes(y = cumulative_frac), size = 1) +
  geom_point(aes(y = cumulative_frac), size = 2) +
  geom_text_repel(
    data = subset(elbow_df, PC %% 10 == 0),
    aes(y = cumulative_frac,
        label = percent(cumulative_frac, accuracy = 1)),
    size = 3
  ) +
  geom_vline(xintercept = elbow_pc, linetype = "dashed", color = "#025E8D") +
  geom_point(
    data = subset(elbow_df, PC == elbow_pc),
    aes(y = variance_frac),
    size = 4, shape = 21, fill = "#025E8D", color = "white"
  ) +
  geom_text_repel(
    data = subset(elbow_df, PC == elbow_pc),
    aes(
      y     = variance_frac,
      label = paste0("Elbow @ PC", elbow_pc)
    ),
    nudge_y = max(elbow_df$variance_frac) * 0.1,
    size     = 4,
    color    = "#025E8D"
  ) +
  # dual y‐axes: left for per‐PC, right for cumulative %
  scale_y_continuous(
    name     = "Per‐PC Variance (fraction)",
    sec.axis = sec_axis(~ .,
                        name   = "Cumulative Variance",
                        labels = percent_format(accuracy = 1))
  ) +
  labs(
    title = "PCA Elbow & Cumulative Variance with Automatic Elbow Detection",
    x     = "Principal Component"
  ) +
  geom_hline(yintercept=0.01, linetype="dashed", color="#025E8D")+
  theme_minimal()

print(p)
```

## PC Heatmap

```{r, fig.asp=3}
PCHeatmap(df, dims = 1:27, cells = 500, balanced = TRUE)
```

```{r}
df$class <- recode_factor(
  df$orig.ident,
  "R3#2_65" = "ctrl_1",
  "TBR2wt"  = "ctrl_2",
  .default  = df$orig.ident,
  .ordered  = TRUE 
)
```

## UMAP

```{r}
Idents(df) <- "class"

df <- RunUMAP(df, reduction="pca", dims=1:25)
DimPlot(df) + NoAxes()
DimPlot(df, split.by="class") + NoAxes()
```

# Clustering

```{r}
DefaultAssay(df)<-"integrated"
df <- FindNeighbors(df, dims = 1:25)
df <- FindClusters(df, resolution = 0.4, graph.name = "integrated_snn", algorithm=4)
DefaultAssay(df)<-"RNA"
df<-JoinLayers(df)
df <- CellCycleScoring(df, s.features = s.genes, g2m.features = g2m.genes)
df <- SCTransform(df, vars.to.regress = c("percent.mt", "S.Score", "G2M.Score"))
```

```{r}
df$celltype<-NULL
# 1) Create a character vector of cell types
celltype_vec <- case_when(
  df$seurat_clusters == 1  ~ "RG+CyP",
  df$seurat_clusters == 2  ~ "Excitatory",
  df$seurat_clusters == 3  ~ "Excitatory",
  df$seurat_clusters == 4  ~ "Stressed_Neurons",
  df$seurat_clusters == 5  ~ "Neuroblasts/IP",
  df$seurat_clusters == 6  ~ "Interneurons",
  df$seurat_clusters == 7  ~ "bRG",
  df$seurat_clusters == 8  ~ "Stressed_progenitors",
  df$seurat_clusters == 9  ~ "ECM-RG",
  df$seurat_clusters == 10 ~ "Interneurons",
  df$seurat_clusters == 11 ~ "EarlyBornNeurons",
  TRUE                      ~ as.character(df$seurat_clusters)
)

# 2) Factor it with your desired order
celltype_fact <- factor(
  celltype_vec,
  levels = c(
    "RG+CyP", "bRG","ECM-RG", "Neuroblasts/IP",
    "EarlyBornNeurons", "Excitatory", "Interneurons",
    "Stressed_progenitors", "Stressed_Neurons"
     
  )
)

# 3) Add it back into your Seurat object’s metadata
df[["celltype"]] <- celltype_fact

# 4) Set identities to celltype
Idents(df) <- "celltype"
```

```{r}
df<-FindSubCluster(df, cluster="Stressed_Neurons", subcluster.name="sub", resolution=0.2, algorithm=4, graph.name="integrated_snn")
Idents(df)<-"sub"
```

```{r}
df$celltype<-NULL
# 1) Create a character vector of cell types
celltype_vec <- case_when(
  df$sub == "Neuroblasts/IP"  ~ "IP",
  df$sub == "RG+CyP"  ~ "RG",
  df$sub == "Stressed_Neurons_1"  ~ "Stressed_Neurons",
  df$sub == "Stressed_Neurons_2"  ~ "Stressed_cells",
  df$sub == "Stressed_progenitors"  ~ "Stressed_cells",
  df$sub == "ECM-RG"  ~ "ECM-P",
  TRUE                      ~ as.character(df$sub)
)

# 2) Factor it with your desired order
celltype_fact <- factor(
  celltype_vec,
  levels = c( "RG", "bRG", "ECM-P", "Stressed_cells", "IP", "EarlyBornNeurons", "Excitatory", "Interneurons", "Stressed_Neurons"
  )
)

# 3) Add it back into your Seurat object’s metadata
df[["celltype"]] <- celltype_fact

# 4) Set identities to celltype
Idents(df) <- "celltype"
```

# DimPlot

```{r}
my.cols <- c(
  "RG"              = "#990000",
  "bRG"             = "#D64528",
  "ECM-P"           = "#FDB267",
  "Stressed_cells"  = "#FEE1CE",
  "IP"              = "#A7D7CF",
  "EarlyBornNeurons"= "#3C9C85",
  "Excitatory"      = "#007C70",
  "Interneurons"    = "#2E5C48",
  "Stressed_Neurons"= "#3D4650"
)

p1<-DimPlot(df, reduction = "umap", group.by  = "celltype" , pt.size=0.1)  & NoAxes() &
  labs(title = NULL) &  scale_color_manual(values = my.cols)
p2<-DimPlot(df, reduction = "umap", group.by  = "celltype" , pt.size=0.1)  & NoAxes() &
  labs(title = NULL) &  scale_color_manual(values = my.cols) & NoLegend()
p1
p2
```

```{r}
# ggsave(
#   filename = "../outputs/Figure5/074_UMAP_w_legend.png",
#   plot     = p1,
#   width    = 8,
#   height   = 8,
#   units    = "cm",
#   dpi      = 600
# )
# 
# # save the no‐legend plot
# ggsave(
#   filename = "../outputs/Figure5/074_UMAP_wo_legend.png",
#   plot     = p2,
#   width    = 12,
#   height   = 12,
#   units    = "cm",
#   dpi      = 600
# )
```

# Cell percentage sample

```{r}
df$celltype <- factor(
  df$celltype,
  levels = names(my.cols)
)

cell_counts <- df@meta.data %>%
  group_by(class, celltype) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(class) %>%
  mutate(percentage = n / sum(n) * 100)

per2 <- ggplot(cell_counts,
               aes(x = class,
                   y = percentage,
                   fill = celltype), position = position_fill(reverse = TRUE) ) +
  geom_bar(
    stat     = "identity",
    position = position_fill(reverse = TRUE)
  ) +
  scale_fill_manual(
    values = my.cols,
    breaks = names(my.cols)       
  ) +
  scale_x_discrete(
    labels = c("ctrl1","ctrl2","mild2","moderate","severe1","severe2")
  ) +
  ylab("Cell‐type proportion") +
  xlab(NULL) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    axis.title.x     = element_text(size = 0),
    axis.title.y     = element_text(size = 0),
    axis.text.x      = element_text(size = 0, angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 0),
    strip.text       = element_text(size = 0),
    plot.title       = element_text(size = 0, face = "bold")
  )

per2

# ggsave(
#   filename = "../outputs/Figure5/074_celltype.png",
#   plot     = per2,
#   width    = 35,
#   height   = 40,
#   units    = "mm",
#   dpi      = 600
# )
```

# DotPlot

```{r, fig.width=16, fig.heigth=4}
Idents(df)<-"celltype"

toplot<-c("NES", "SOX2", "PAX6", "FOXG1","VIM", "SLC1A3","GLI3",  "TTYH1", "RFX4", "SHROOM3",
          "MKI67", "POLQ", "CENPK", "TOP2A", "CENPF", "ASPM",
          "PTN", "PTPRZ1", "HOPX", "PEA15", "ITGB5",
          "COL1A1","COL1A2","COL3A1","PDGFRA","DCN","POSTN",
          "EOMES", "HES6", "ST18", "IGFBPL1", "NEUROG2","DCX",
          "MAP2", "TUBB3",
           "RORB", "SATB2", "NEUROD6", "SLC24A2", "GRIA2", "CACNA1E",
          "TBR1",  "BCL11B","GRIN2B","RALYL","ARPP21",
          "GAD1","GAD2","LHX1", "DLX5", "DLX6", 
          "SNHG8","GAS5","RPL36AL","HSPB1","NME2", "HSPA5", "DDIT3", "ATF4", "ATF6", "XBP1", "HSP90B1")
toplot <- unique(toplot) 

p11 <- DotPlot(df, features = toplot, cols = diverging_palette_4(5))
filtered_data <- p11$data %>%
  group_by(features.plot) %>%
  filter(any(pct.exp >= 0)) %>%
  ungroup()
expression_matrix <- filtered_data %>%
  dplyr::select(features.plot, id, avg.exp.scaled) %>%
  tidyr::pivot_wider(names_from = id, values_from = avg.exp.scaled, values_fill = 0) %>%
  tibble::column_to_rownames("features.plot")
gene_order <- hclust(dist(expression_matrix))$order
sorted_genes <- rownames(expression_matrix)[gene_order]
p11 <- DotPlot(df, features = toplot, cols = diverging_palette_4(5)) +
  scale_color_gradientn(
    colors = diverging_palette_4(5), 
    limits = c(-max(abs(filtered_data$avg.exp.scaled)), max(abs(filtered_data$avg.exp.scaled)))
  ) +
  #coord_flip() +
  #labs(title = "Filtered DotPlot: Genes clustered by similarity") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1,size=8)) +
  theme(axis.text.x=element_text(angle=45, hjust=1, size=8))


p11_final <- p11 +
  labs(x = NULL, y = NULL) +
  scale_size_area(max_size = 3) +
  theme(
    legend.position="none",
    axis.text.x      = element_text(size = 6, angle = 45, hjust = 1),
    axis.text.y      = element_blank(),
    axis.title.x     = element_blank(),
    axis.title.y     = element_blank(),
    plot.title       = element_blank()
  )

# 2) Save it at 105×45 mm, 600 dpi
# ggsave(
#   filename = "../outputs/Figure5/074_dotplot_new.png",
#   plot     = p11_final,
#   width    = 190,
#   height   = 60,
#   units    = "mm",
#   dpi      = 600
# )
```

# GO enrichment dotplot severe

## All neurons (excluding stressed)

```{r}
Idents(df)<-"celltype"
dfb<-df
df <- subset(df, idents = c("EarlyBornNeurons", "Excitatory", "Interneurons"))
DefaultAssay(df)<-"SCT"
df$severity<-NULL
# 1) Create a character vector of cell types
celltype_vec <- case_when(
  df$class == "ctrl_1"  ~ "ctrl",
  df$class == "ctrl_2"  ~ "ctrl",
  df$class == "mild_1"  ~ "mild",
  df$class == "mild_2"  ~ "mild",
  df$class == "mod_2"  ~ "moderate",
  df$class == "sev_1"  ~ "severe",
  df$class == "sev_2"  ~ "severe",
  
  TRUE                      ~ as.character(df$seurat_clusters)
)

# 2) Factor it with your desired order
celltype_fact <- factor(
  celltype_vec,
  levels = c("ctrl", "mild", "moderate", "severe"
  )
)

# 3) Add it back into your Seurat object’s metadata
df[["severity"]] <- celltype_fact

# 4) Set identities to celltype
Idents(df) <- "severity"
```

TOP 500 DEGs (250 up and 250 down)

```{r, warning=FALSE, message=FALSE, results='hide'}
library(purrr)

get_markers <- function(object, ident.1, ident.2 = "ctrl",  assay = "SCT", min.pct = 0.25, log2.threshold = 0.5,  n.sig = 250, n.top = 10) {
  
  res <- FindMarkers(
    object,
    ident.1      = ident.1,
    ident.2      = ident.2,
    assay        = assay,
    min.pct      = min.pct,
    log2.threshold = log2.threshold
  ) %>%
    tibble::rownames_to_column("gene")
  
  sig <- res %>%
    filter(p_val_adj < 0.05, abs(avg_log2FC) > log2.threshold) %>%
    arrange(desc(avg_log2FC)) %>%
    slice_head(n = n.sig) %>%
    bind_rows(
      res %>%
        filter(p_val_adj < 0.05, abs(avg_log2FC) > log2.threshold) %>%
        arrange(avg_log2FC) %>%
        slice_head(n = n.sig)
    )
  
  top10 <- sig %>%
    slice_head(n = n.top)  
  
  list(
    sig   = sig,
    top10 = top10
  )
}

severity_levels <- c("mild", "moderate", "severe")

marker_results <- map(
  severity_levels,
  ~ get_markers(df, ident.1 = .x, ident.2 = "ctrl")
)

names(marker_results) <- severity_levels

severe_vs_ctrl_sig   <- marker_results[["severe"]]$sig
severe_vs_ctrl_top10 <- marker_results[["severe"]]$top10
```

### Volcano plot (stressed and apoptotic markers)

```{r}
p_volcano<-make_volcano_v2(df, ident.1=c("severe"), ident.2=c("ctrl"), assay="SCT", highlight_genes = c("ATF4", "ATF6", "XBP1", "DDIT3", "CHOP", "HSPA5", "GRP78", "EIF2AK3", "PERK",
  "ERN1", "IRE1", "HERPUD1", "DNAJB9", "EDEM1", "MANF", "PDIA4", "HSP90B1", "BAX", "BAK1", "BCL2", "BCL2L1", "MCL1", "PUMA", "NOXA", "BID", "CASP3",  "CASP7", "CASP8", "CASP9", "FAS", "FADD", "TNFRSF10A", "TNFRSF10B", "TP53"))
```
```{r}
# p_volcano_8pt <- p_volcano +
#   theme(
#     text             = element_text(size = 8),
#     axis.title       = element_text(size = 8),
#     axis.text        = element_text(size = 8),
#     legend.text      = element_text(size = 8),
#     legend.title     = element_text(size = 8),
#     plot.title       = element_text(size = 8),
#     plot.subtitle    = element_text(size = 8),
#     plot.caption     = element_text(size = 8)
#   )
# 
# ggsave(
#   filename = "../outputs/Figure5/074_volcano_severe_vs_ctrl.png",
#   plot     = p_volcano_8pt,
#   width    = 180,    # mm
#   height   = 100,    # mm
#   units    = "mm",
#   dpi      = 600
# )
```

```{r}
severe_entrez <- bitr(severe_vs_ctrl_sig$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

go_severe <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "BP",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)

go_severe_MF <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "MF",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)


go_severe_CC <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "CC",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)


kegg_severe <- enrichKEGG(
  gene = severe_entrez$ENTREZID,
  organism = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)
```

```{r}
severe_df_bp<-as.data.frame(go_severe@result)
severe_df_mf<-as.data.frame(go_severe_MF@result)
severe_df_cc<-as.data.frame(go_severe_CC@result)
severe_kegg_df<-as.data.frame(kegg_severe@result)

severe_df_bp <- severe_df_bp %>% filter(p.adjust<0.05)
severe_df_mf <- severe_df_mf %>% filter(p.adjust<0.05)
severe_df_cc <- severe_df_cc %>% filter(p.adjust<0.05)
severe_kegg_df<-severe_kegg_df %>% filter(p.adjust<0.05)
```

### GO ENRICHMENT STRESS

```{r, fig.width=8}
df_list <- list(
  severe_df_bp, severe_df_mf, severe_df_cc, severe_kegg_df
)

conditions <- c("Severe", "Severe", "Severe", "Severe")


go_tests <- c("BP", "MF", "CC", "KEGG")

keywords <- c("endoplasmic", "unfolded", "misfolded", "fold", "protein binding", "hypoxia", "HIF", "apopto", "oxygen species", "ubiquitin")
```

```{r}
keep_idx   <- which(conditions == "Severe")
df_list    <- df_list[keep_idx]
conditions <- conditions[keep_idx]
go_tests   <- go_tests[keep_idx]

stopifnot(length(df_list) == length(conditions),
          length(df_list) == length(go_tests))

pattern <- paste(keywords, collapse = "|")

processed_list <- mapply(function(df, cond, go_cat) {
  df %>%
    filter(grepl(pattern, Description, ignore.case = TRUE)) %>%
    transmute(
      Description,
      enrichment  = FoldEnrichment,
      fill_metric = -log10(p.adjust),
      GO_test     = go_cat
    )
}, df_list, conditions, go_tests, SIMPLIFY = FALSE)

combined_df <- bind_rows(processed_list) %>%
  arrange(GO_test, Description) %>%
  mutate(
    Description = factor(Description, levels = unique(Description))
  )

p_bars_clean <- ggplot(combined_df,
                       aes(x = enrichment, y = Description, fill = fill_metric)) +
  geom_col(color = "black") +
  scale_fill_gradient(name = NULL) +
  labs(x = "Fold Enrichment", y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y     = element_text(size = 8),
    axis.text.x     = element_text(size = 8, color = "black"),
    axis.ticks.x    = element_line(color = "black"),
    panel.grid.major.x = element_line(linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_blank(),
    legend.position = "none",
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    plot.caption    = element_blank()
  )

print(p_bars_clean)

# ggsave("../outputs/Figure5/074_GO_barplot_STRESS.png", plot = p_bars_clean,
#        width  = 70*2,   
#        height = 90*2  , 
#        units  = "mm",
#        dpi    = 600) 
```

### GO Enrichment cytoskeleton + WNT + Synapse

```{r, fig.width=8}
df_list <- list(
  severe_df_bp, severe_df_mf, severe_df_cc,severe_kegg_df
)

conditions <- c("Severe", "Severe", "Severe", "Severe")

go_tests <- c("BP", "MF", "CC", "KEGG")

keywords <- c("microtubule", "adhesion", "WNT", "tubulin", "cytoskelet", "synapse", "axon")

compare_go_manuscript <- function(df_list, keywords, conditions, go_tests) {
  keep_idx <- which(conditions == "Severe")
  df_list     <- df_list[keep_idx]
  conditions  <- conditions[keep_idx]
  go_tests    <- go_tests[keep_idx]
  
  
  if(length(df_list) != length(conditions) || length(df_list) != length(go_tests)) {
    stop("df_list, conditions, and go_tests must have the same length.")
  }
  

  pattern <- paste(keywords, collapse = "|")
  

  filter_and_tag <- function(df, condition, go_test) {
    df %>%
      filter(grepl(pattern, Description, ignore.case = TRUE)) %>%
      mutate(Condition = condition,
             GO_test = go_test,
             enrichment = -log10(p.adjust))
  }
  
  
  processed_list <- mapply(function(df, cond, go_cat) {
    filter_and_tag(df, cond, go_cat)
  }, df_list, conditions, go_tests, SIMPLIFY = FALSE)
  

  combined_df <- bind_rows(processed_list) %>%
    arrange(GO_test, Description) %>%
    mutate(Description = factor(Description, levels = unique(Description)))
  
  annotation_df <- combined_df %>%
    distinct(Description, GO_test) %>%
    mutate(GO_test = factor(GO_test, levels = rev(c("KEGG", "CC", "MF", "BP"))))
  
 
  p_annotation <- ggplot(annotation_df, aes(x = 1, y = Description, fill = GO_test)) +
    geom_tile() +
    scale_x_continuous(expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = "GO Category") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          panel.grid = element_blank(),
          legend.position = "none") +
    scale_fill_manual(values = c("BP" = "#025e8d", 
                                 "MF" = "#be1818", 
                                 "CC" = "#00a69d", 
                                 "KEGG"="#ffcc00"))
  

  combined_df <- combined_df %>%
    mutate(fill_metric = -log10(p.adjust))
  
  p_main <- ggplot(combined_df, aes(x = Condition, y = Description)) +
    geom_point(aes(size = enrichment, fill = fill_metric), shape = 21, color = "black") +
    scale_fill_gradient(low = "#8FD3F4", high = "#03396c") +
    labs(x = "", y = "", size = "Fold Enrichment", fill = "-log10(p.adjust)") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
  

  combined_plot <- p_annotation + p_main +
    patchwork::plot_layout(widths = c(0.1, 4))
  
  return(combined_plot)
}

p<-compare_go_manuscript(df_list, keywords, conditions, go_tests)
p
```
```{r}
compare_go_manuscript_bar <- function(df_list, keywords, conditions, go_tests) {

  keep_idx   <- which(conditions == "Severe")
  df_list    <- df_list[keep_idx]
  conditions <- conditions[keep_idx]
  go_tests   <- go_tests[keep_idx]
  

  if(length(df_list) != length(conditions) || length(df_list) != length(go_tests)) {
    stop("Inputs must have the same length")
  }
  

  pattern <- paste(keywords, collapse = "|")
  

  processed_list <- mapply(function(df, cond, go_cat) {
    df %>%
      filter(grepl(pattern, Description, ignore.case = TRUE)) %>%
      mutate(
        Condition   = cond,
        GO_test     = go_cat,
        enrichment  = FoldEnrichment,      
        fill_metric = -log10(p.adjust)
      )
  }, df_list, conditions, go_tests, SIMPLIFY = FALSE)
  
  combined_df <- bind_rows(processed_list) %>%
    arrange(GO_test, Description) %>%
    mutate(
     
      Description = factor(Description, levels = unique(Description)),
      GO_test     = factor(GO_test, levels = rev(c("KEGG","CC","MF","BP")))
    )
  

  p_annotation <- ggplot(
    combined_df %>% distinct(Description, GO_test),
    aes(x = 1, y = Description, fill = GO_test)
  ) +
    geom_tile() +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_manual(values = c(
      "BP"   = "#025e8d",
      "MF"   = "#be1818",
      "CC"   = "#00a69d",
      "KEGG" = "#ffcc00"
    )) +
    theme_minimal() +
    theme(
      axis.text.x      = element_blank(),        
      axis.ticks.x     = element_blank(),
      panel.grid       = element_blank(),
      legend.position  = "none"
 
    ) +
    labs(x = NULL, y = NULL)
  

  p_bars <- ggplot(
    combined_df,
    aes(x = enrichment, y = Description, fill = fill_metric)
  ) +
    geom_col(color = "black") +
    scale_fill_gradient(
      low  = "#8FD3F4",
      high = "#03396c",
      name = "-log10(p.adjust)"
    ) +
    labs(x = "Fold Enrichment", y = NULL) +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
  

  (p_annotation + p_bars) +
    plot_layout(widths = c(0.1, 1))
}


p_bar <- compare_go_manuscript_bar(df_list, keywords, conditions, go_tests)
print(p_bar)
```

```{r}
keep_idx   <- which(conditions == "Severe")
df_list    <- df_list[keep_idx]
conditions <- conditions[keep_idx]
go_tests   <- go_tests[keep_idx]

stopifnot(length(df_list) == length(conditions),
          length(df_list) == length(go_tests))


pattern <- paste(keywords, collapse = "|")


processed_list <- mapply(function(df, cond, go_cat) {
  df %>%
    filter(grepl(pattern, Description, ignore.case = TRUE)) %>%
    transmute(
      Description,
      enrichment  = FoldEnrichment,
      fill_metric = -log10(p.adjust),
      GO_test     = go_cat
    )
}, df_list, conditions, go_tests, SIMPLIFY = FALSE)


combined_df <- bind_rows(processed_list) %>%
  arrange(GO_test, Description) %>%
  mutate(
    Description = factor(Description, levels = unique(Description))
  )


p_bars_clean <- ggplot(combined_df,
                       aes(x = enrichment, y = Description, fill = fill_metric)) +
  geom_col(color = "black") +
  scale_fill_gradient(name = NULL) +   
  labs(x = "Fold Enrichment", y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y     = element_text(size = 8),
    axis.text.x     = element_text(size = 8, color = "black"),
    axis.ticks.x    = element_line(color = "black"),
    panel.grid.major.x = element_line(linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_blank(),
    legend.position = "none",
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    plot.caption    = element_blank()
  )

print(p_bars_clean)

# ggsave("../outputs/Figure5/GO_barplot_70x80mm.png", plot = p_bars_clean,
#        width  = 70*2,   
#        height = 90*2  ,  
#        units  = "mm",
#        dpi    = 600)  
```

##  Stressed neurons

```{r}
Idents(dfb)<-"celltype"
df <- subset(dfb, idents = c("Stressed_Neurons"))
DefaultAssay(df)<-"SCT"
Idents(df)<-"severity"
```

TOP 500 DEGs (250 up and 250 down)

```{r, warning=FALSE, message=FALSE, results='hide'}
severity_levels <- c("severe")

marker_results <- map(
  severity_levels,
  ~ get_markers(df, ident.1 = .x, ident.2 = "ctrl")
)
names(marker_results) <- severity_levels

severe_vs_ctrl_sig   <- marker_results[["severe"]]$sig
severe_vs_ctrl_top10 <- marker_results[["severe"]]$top10
```

### Volcano plot (stressed and apoptotic markers)

```{r}
P_volcano2<-make_volcano_v2(df, ident.1=c("severe"), ident.2=c("ctrl"), assay="SCT", highlight_genes = c("ATF4", "ATF6", "XBP1", "DDIT3", "CHOP", "HSPA5", "GRP78", "EIF2AK3", "PERK",
  "ERN1", "IRE1", "HERPUD1", "DNAJB9", "EDEM1", "MANF", "PDIA4", "HSP90B1", "BAX", "BAK1", "BCL2", "BCL2L1", "MCL1", "PUMA", "NOXA", "BID", "CASP3",  "CASP7", "CASP8", "CASP9", "FAS", "FADD", "TNFRSF10A", "TNFRSF10B", "TP53"))
P_volcano2
```

```{r}
# p_volcano_8pt <- P_volcano2 +
#   theme(
#     text             = element_text(size = 8),
#     axis.title       = element_text(size = 8),
#     axis.text        = element_text(size = 8),
#     legend.text      = element_text(size = 8),
#     legend.title     = element_text(size = 8),
#     plot.title       = element_text(size = 8),
#     plot.subtitle    = element_text(size = 8),
#     plot.caption     = element_text(size = 8)
#   )
# 
# ggsave(
#   filename = "../outputs/Figure5/074_volcano_severe_vs_ctrl_stressed_neurons.png",
#   plot     = p_volcano_8pt,
#   width    = 180,    # mm
#   height   = 100,    # mm
#   units    = "mm",
#   dpi      = 600
# )
```


```{r}
severe_entrez <- bitr(severe_vs_ctrl_sig$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

go_severe <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "BP",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)

go_severe_MF <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "MF",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)

go_severe_CC <- enrichGO(
  gene = severe_entrez$ENTREZID,     
  OrgDb = org.Hs.eg.db,            
  keyType = "ENTREZID",           
  ont = "CC",                     
  pAdjustMethod = "BH",           
  pvalueCutoff = 0.05,            
  qvalueCutoff = 0.05,            
  readable = TRUE                 
)

kegg_severe <- enrichKEGG(
  gene = severe_entrez$ENTREZID,
  organism = "hsa",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05
)
```

```{r}
severe_df_bp<-as.data.frame(go_severe@result)
severe_df_mf<-as.data.frame(go_severe_MF@result)
severe_df_cc<-as.data.frame(go_severe_CC@result)
severe_kegg_df<-as.data.frame(kegg_severe@result)

severe_df_bp <- severe_df_bp %>% filter(p.adjust<0.05)
severe_df_mf <- severe_df_mf %>% filter(p.adjust<0.05)
severe_df_cc <- severe_df_cc %>% filter(p.adjust<0.05)
severe_kegg_df<-severe_kegg_df %>% filter(p.adjust<0.05)
```

### GO ENRICHMENT STRESS

```{r, fig.width=8}
df_list <- list(
  mild_df_bp, mild_df_mf, mild_df_cc,
  moderate_df_bp, moderate_df_mf, moderate_df_cc,
  severe_df_bp, severe_df_mf, severe_df_cc,
  mild_kegg_df, moderate_kegg_df, severe_kegg_df
)

conditions <- c("Severe", "Severe", "Severe","Severe")

go_tests <- c("BP", "MF", "CC", "KEGG")

keywords <- c("endoplasmic", "unfolded", "misfolded", "fold", "protein binding", "hypoxia", "HIF", "apopto", "oxygen species", "ubiquitin")

p<-compare_go_manuscript(df_list, keywords, conditions, go_tests)
p
```
```{r}

p_bar <- compare_go_manuscript_bar(df_list, keywords, conditions, go_tests)
print(p_bar)
```

```{r}
keep_idx   <- which(conditions == "Severe")
df_list    <- df_list[keep_idx]
conditions <- conditions[keep_idx]
go_tests   <- go_tests[keep_idx]

stopifnot(length(df_list) == length(conditions),
          length(df_list) == length(go_tests))

pattern <- paste(keywords, collapse = "|")

processed_list <- mapply(function(df, cond, go_cat) {
  df %>%
    filter(grepl(pattern, Description, ignore.case = TRUE)) %>%
    transmute(
      Description,
      enrichment  = FoldEnrichment,
      fill_metric = -log10(p.adjust),
      GO_test     = go_cat
    )
}, df_list, conditions, go_tests, SIMPLIFY = FALSE)

combined_df <- bind_rows(processed_list) %>%
  arrange(GO_test, Description) %>%
  mutate(
    Description = factor(Description, levels = unique(Description))
  )


p_bars_clean <- ggplot(combined_df,
                       aes(x = enrichment, y = Description, fill = fill_metric)) +
  geom_col(color = "black") +
  scale_fill_gradient(name = NULL) +  
  labs(x = "Fold Enrichment", y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y     = element_text(size = 8),
    axis.text.x     = element_text(size = 8, color = "black"),
    axis.ticks.x    = element_line(color = "black"),
    panel.grid.major.x = element_line(linetype = "dotted"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.title.x    = element_text(size = 8),
    axis.title.y    = element_blank(),
    legend.position = "none",
    plot.title      = element_blank(),
    plot.subtitle   = element_blank(),
    plot.caption    = element_blank()
  )

print(p_bars_clean)

# ggsave("path", plot = p_bars_clean,
#        width  = 70*2,    # mm
#        height = 90*2  ,    # mm
#        units  = "mm",
#        dpi    = 600)
```

```{r}
sessionInfo()
```

