---
title: "QC & filtering"
author: "Anne Hoffrichter, Eric Poisel, Lea Zillich"
date: "2023/02/23"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80),tidy=TRUE, fig.asp=0.5, fig.width=12, warning = FALSE)
knitr::opts_knit$set(root.dir = "..")
getwd()
```

```{r loadLibraries, message=FALSE}
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(Matrix)
library(dplyr)
options(stringsAsFactors = FALSE)
```

- Specify path to raw data 
```{r definepathtodata, message=FALSE}
barcodes <- "path/to/barcodes.tsv.gz"
features <- "path/to/features.tsv.gz"
matrix <- "path/to/matrix.mtx.gz"
```

### Load data
Feature barcode matrix is loaded and transformed into the SCE format.
```{r loadData}
# Import raw data files
cellbarcodes <- read.table(barcodes)
genenames <- read.table(features)
molecules <- readMM(matrix)

rownames(molecules) <- genenames[,1]
colnames(molecules) <- cellbarcodes[,1]
umi <- SingleCellExperiment(assays=list(counts=as.matrix(molecules)))
```

### Start QC & Filtering
- Remove genes that are not expressed in any cell
```{r KeepOnlyGenesExpressedInAtLeastOneCell}
keep_feature <- rowSums(counts(umi) > 0) > 0
umi <- umi[keep_feature, ]
```

- Define control features for  MT genes
```{r DefineMTgenes}
is.spike <- rownames(umi) %in% 
 c("ENSG00000198899", "ENSG00000198727", "ENSG00000198888",
   "ENSG00000198886", "ENSG00000212907", "ENSG00000198786",
   "ENSG00000198695", "ENSG00000198712", "ENSG00000198804",
   "ENSG00000198763", "ENSG00000228253", "ENSG00000198938",
   "ENSG00000198840")
umi <- splitAltExps(umi, ifelse(is.spike, "MT", "gene"))
```

- Calculate quality metrics
```{r qualityMetricsMT, warning=FALSE}
umi2 <- perCellQCMetrics(umi, use_altexps = TRUE)
```

### Cell QC   
Set the count_thres value so that a normal distribution is observed in the middle and right part of the histogram i.e. exclude cells on the left of the histogram
```{r HistTotalCounts}
count_thres <- 2800

hist(
  umi2$total,
  breaks = 400,
  xlim=range(1:30000)
)
abline(v = count_thres, col = "red")
```

```{r FilterTotalCounts}
filter_by_total_counts <- (umi2$total > count_thres)
table(filter_by_total_counts)
```

Set the feature threshold so that a normal distribution is observed in the middle and right part of the histogram i.e. exclude cells on the left of the histogram
```{r HistDetectedGenes}
feature_thres <- 1400

hist(
  umi2$detected,
  breaks = 100
)
abline(v = feature_thres, col = "red")
```

```{r FilterExprFeatures}
filter_by_expr_features <- (umi2$detected > feature_thres)
table(filter_by_expr_features)
```

```{r PlotMTgenes}
plot(
x = umi2$detected,
y = umi2$altexps_MT_percent
)
```

Create filters for removing cells with high expression of mitochondrial genes (exclusion if mt counts >10% of total counts in a cell is generally a good choice, only change if necessary)

```{r FilterMTgenes}
mt_thres <- 10 

filter_by_MT <- umi2$altexps_MT_percent < mt_thres
table(filter_by_MT)
```

Defining a cell filter based on the previous analysis.   
Number of cells that will be removed by the filters:

```{r AllFilters}
umi2$use <- (
  # sufficient features (genes)
  filter_by_expr_features &
  # sufficient molecules counted
  filter_by_total_counts &
  # remove cells with unusual number of reads in MT genes
  filter_by_MT
)
table(umi2$use)
```

```{r FilterParameters}
param <- list()
param$total_counts <- count_thres
param$expr_features <- feature_thres
param$MT <- mt_thres

# Export threshold values to a file 
write.table(param, "/path/to/001_filter_param_C1.txt", quote = F)
```

```{r sessioninfo}
sessionInfo()
```