---
title: "Figures & Analyses"
author: "Lea Zillich, Anne Hoffrichter, Eric Poisel"
date: "2022/02/17"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80),tidy=TRUE, fig.asp=0.5, fig.width=12, warning = FALSE)
```

```{r loadLibraries, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(cowplot)
library(openxlsx)
library(dplyr)
library(readxl)
#library(msigdb)
library(enrichR)
library(msigdbr)
library(org.Hs.eg.db)
library(tidyverse)
library(patchwork)
library(GO.db)
library(ggpubr)
library(pheatmap)
library(VennDiagram)
set.seed(42)
```

- Import integrated object
```{r importObject}
sc <- readRDS(file = "/path/to/4_SeuratObjects/LIS1_integrated.rds")
#sc <- readRDS(file = "path/to/4_SeuratObjects/LIS1_integrated.rds")
```

```{r normalize data}
# Set Idents to clusters names
Idents(sc) <- "cell_types"

# Normalize and scale RNA assay
sc <- NormalizeData(sc, assay = "RNA")
sc <- FindVariableFeatures(sc, selection.method = "vst", assay = "RNA")
sc <- ScaleData(sc, features = rownames(sc), assay = "RNA")
```


- Figure 2B & Sup Figure S2A
```{r DimPLots data}
p1 <- DimPlot(sc, reduction = "umap",  cols = c("#C7361B","#F7AB64", "#313C48","#880000","#006960","#FCD8C1","#428E77","#778CA1","#BFE1D7","#21473C"))

ggsave("/path/to/figures/DimPlot.png", p1, width= 7, height = 5, dpi=600)

p2 <- DimPlot(sc, reduction = "umap",  split.by = "replicate",cols = c("#C7361B","#F7AB64", "#313C48","#880000","#006960","#FCD8C1","#428E77","#778CA1","#BFE1D7","#21473C"),ncol=4)


#ggsave("/path/to/figures/DimPlot_by_replicate.pdf", p2, width= 9, height = 6, dpi=600)

p1
p2
```


```{r FeaturePLots NE}

p3 <- FeaturePlot(sc, features=c("ZBTB16", "MMRN1","NR2F1","BTBD17"))


p3
```

- Figure 1H

```{r celltypes stacked}
samples <- c(rep("Ctrl_C1" , 10),rep("Ctrl_C7" , 10) , rep("Mild_P1" , 10), rep("Mild_P2" , 10), rep("Moderate_P3" , 10), rep("Moderate_P4" , 10), rep("Severe_P5" , 10), rep("Severe_P6" , 10))

celltypes <- rep(names(table(sc$cell_types)) , 8)
prop <- table(sc$replicate,sc$cell_types)

value <- c(unname(prop[1,]),unname(prop[2,]),unname(prop[3,]),unname(prop[4,]),unname(prop[5,]),unname(prop[6,]),unname(prop[7,]),unname(prop[8,]))

cell_perc <- data.frame(samples,celltypes,value)
cell_perc$celltypes <- factor(cell_perc$celltypes, levels =c("IP","vFB-N","dFB-N","MB-N","IN","G","t-RG","RG","CyP","NE"))
     
cell_perc$samples <- factor(cell_perc$samples)

# Stacked + percent
per <- ggplot(cell_perc, aes(fill=celltypes, y=value, x=samples)) + 
    geom_bar(position="fill", stat="identity")+scale_fill_manual(values=c("#21473C","#428E77","#006960","#BFE1D7","#778CA1","#313C48", "#880000","#C7361B","#F7AB64","#FCD8C1"))+ylab("celltype proportion")+xlab(NULL)+
  theme_void()+theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), axis.text.y = element_text(), axis.title.y = element_text(angle = 90,vjust=1))
ggsave(plot = per,filename ="/path/to/figures/celltype_percentage_revision.pdf", width = 3, height=5)



```



## Supplementary Figures 3B

```{r sut-RG fig 3b 3d, warning=FALSE, message=FALSE, results='hide'}
condition <- c(rep("Ctrl" , 10),rep("Mild" , 10),rep("Moderate" , 10), rep("Severe" , 10))

celltypes <- rep(names(table(sc$cell_types)) , 4)
prop <- table(sc$condition,sc$cell_types)

value <- c(unname(prop[1,]),unname(prop[2,]),unname(prop[3,]),unname(prop[4,]))

cell_perc <- data.frame(condition,celltypes,value)
cell_perc$celltypes <- factor(cell_perc$celltypes, levels =c("IP","vFB-N","dFB-N","MB-N","IN","G","t-RG","RG","CyP","NE"))
     
cell_perc$condition<- factor(cell_perc$condition)


cell_perc$percent <- c(cell_perc$value[cell_perc$condition=="Ctrl"]/sum(cell_perc$value[cell_perc$condition=="Ctrl"]),
                       cell_perc$value[cell_perc$condition=="Mild"]/sum(cell_perc$value[cell_perc$condition=="Mild"]),
                       cell_perc$value[cell_perc$condition=="Moderate"]/sum(cell_perc$value[cell_perc$condition=="Moderate"]),
                       cell_perc$value[cell_perc$condition=="Severe"]/sum(cell_perc$value[cell_perc$condition=="Severe"]))

cell_prog <- data.frame(condition=rep(c("Ctrl","Mild","Moderate","Severe"),2),
                        type=c(rep("Progenitor",4),rep("Differentiated Cells",4)),
                        perc=c(sum(cell_perc$percent[cell_perc$condition=="Ctrl"& (cell_perc$celltypes=="RG"|cell_perc$celltypes=="t-RG"|cell_perc$celltypes=="NE")]),sum(cell_perc$percent[cell_perc$condition=="Mild"& (cell_perc$celltypes=="RG"|cell_perc$celltypes=="t-RG")]), sum(cell_perc$percent[cell_perc$condition=="Moderate"&  (cell_perc$celltypes=="RG" |cell_perc$celltypes=="t-RG"|cell_perc$celltypes=="NE")]), sum(cell_perc$percent[cell_perc$condition =="Severe"& (cell_perc$celltypes=="RG"|cell_perc$celltypes=="t-RG"|cell_perc$celltypes=="NE")]), sum(cell_perc$percent[cell_perc$condition=="Ctrl"& (cell_perc$celltypes =="dFB-N" |cell_perc$celltypes=="vFB-N"|cell_perc$celltypes=="MB-N"|cell_perc$celltypes=="IN"|cell_perc$celltypes=="G")]),sum(cell_perc$percent[cell_perc$condition=="Mild"& (cell_perc$celltypes=="dFB-N" |cell_perc$celltypes=="vFB-N"|cell_perc$celltypes=="MB-N"|cell_perc$celltypes=="IN"|cell_perc$celltypes=="G")]), sum(cell_perc$percent[cell_perc$condition=="Moderate"& (cell_perc$celltypes=="dFB-N" |cell_perc$celltypes=="vFB-N"|cell_perc$celltypes=="MB-N"|cell_perc$celltypes=="IN"|cell_perc$celltypes=="G")]),sum(cell_perc$percent[cell_perc$condition=="Severe"& (cell_perc$celltypes=="dFB-N" |cell_perc$celltypes=="vFB-N"|cell_perc$celltypes=="MB-N"|cell_perc$celltypes=="IN"|cell_perc$celltypes=="G")])))
cell_prog$type <- factor(cell_prog$type, levels = c("Progenitor","Differentiated Cells"))

df_bar_prog <- cell_perc[cell_perc$celltypes=="IP"|cell_perc$celltypes=="t-RG"|cell_perc$celltypes=="RG"|cell_perc$celltypes=="CyP"|cell_perc$celltypes=="NE",]

bar_prog <- ggplot(df_bar_prog, aes(x=celltypes, y=percent, fill=condition, color=condition)) + ylim(0,1) +
  geom_bar(stat="identity", color="black",position=position_dodge())+theme_classic()+
  labs(y="cells (%)") + scale_fill_manual(values=c()) + theme(text = element_text(size=18),axis.title.x=element_blank(), axis.text.y=element_text(size=18), legend.text = element_text(size=14), axis.ticks.x = element_blank(),legend.position = "bottom") + scale_fill_manual(values=c("gray93","gray68","gray43","gray23"), labels=c("control","mild","moderate","severe"))

#ggsave(ggarrange(bar_cell,bar_prog,widths = c(0.4,0.6),common.legend = T),filename ="/path/to/figures/SupFig3B_barplot_progenitor.pdf", width = 10, height=7)

ggsave(bar_prog,filename ="/path/to/figures/barplot_progenitor.pdf", width = 10, height=7)

```

- Supplement Figure S3A
```{r dotplot markers}
genes <- c("TOP2A","TPX2","MKI67","CTNNB1","CDH2", "FABP7", "HES1","HES4","HES5", "SOX2", "SOX4", "VIM", "NES", "ANXA5", "PTPRZ1", "FABP5","C1orf61", "OTX2", "PAX6", "EMX2", "SFRP1", "NEUROD1","EOMES",  "ELAVL4", "TBR1", "STMN1", "DCX", "MAP2", "DLX1", "DLX2", "SLC32A1", "ARX", "SST", "ROBO2", "LMX1B","EN1", "EN2", "MOXD1", "S100B")

sc@active.ident <- factor(sc@active.ident, 
                            levels=c("NE",
                                     "CyP",
                                     "RG",
                                     "t-RG",
                                     "IP",
                                     "dFB-N", 
                                     "vFB-N", 
                                     "IN", 
                                     "MB-N", 
                                     "G"))


p3 <- DotPlot(object = sc, features = genes,assay ="RNA", dot.scale = 5, cols = c("lightgrey","purple4")) + 
  theme(axis.text.x = element_text(angle = 90)) + 
  theme(axis.text.y = element_text(size=12)) + 
  theme(axis.text.x = element_text(size=12)) + 
  theme(text=element_text(size = 12)) + 
  theme(aspect.ratio =0.5)

#ggsave("/path/to/figures/DotPlot.png", p3, width= 10, height = 5, dpi=600)
```



## Differential expression analysis

```{r DEanalysis, warning=FALSE, message=FALSE, results='hide'}

sc@meta.data$progenitor <- ifelse(sc@meta.data$cell_types==c("RG", "NE", "t-RG"),"progenitor",NA)

sc@meta.data$DE_progenitor <- paste0(sc@meta.data$condition,"_",sc@meta.data$progenitor)
Idents(sc) <- "DE_progenitor"

DefaultAssay(sc) <- "RNA"

for (i in c("Mild","Moderate","Severe")){

markers_cl <- FindMarkers(sc,ident.1 = paste0(i,"_progenitor"), ident.2 = "Ctrl_progenitor", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
head(markers_cl,100)

}


#compare moderate and mild
markers_cl <- FindMarkers(sc,ident.1 = "Moderate_progenitor", ident.2 = "Mild_progenitor", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
write.csv(markers_cl, "/path/to/DEresults/progenitor_moderate_vs_mild.csv")

#compare severe and mild
markers_cl <- FindMarkers(sc,ident.1 = "Severe_progenitor", ident.2 = "Mild_progenitor", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
write.csv(markers_cl, "/path/to/DEresults/progenitor_severe_vs_mild.csv")

#compare severe and moderate
markers_cl <- FindMarkers(sc,ident.1 = "Severe_progenitor", ident.2 = "Moderate_progenitor", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
write.csv(markers_cl, "/path/to/DEresults/progenitor_severe_vs_moderate.csv")

```


```{r DEanalysis NE marker, warning=FALSE, message=FALSE, results='hide'}

Idents(sc) <- "cell_types"

DefaultAssay(sc) <- "RNA"

markers_rg <- FindMarkers(sc,ident.1 = "NE", ident.2 = "RG", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
head(markers_rg,100)

write.csv(markers_rg, "path/to/DEresults/NE_markers_vs_RG.csv", row.names = T)

```


# Heatmap of DE genes across clusters - Figure 2E

```{r hm_GEX,warning=FALSE, message=FALSE, results='hide'}
# Import DE results - log2FC 0.5 cutoff and padj 0.05, only celltypes with more than 100 cells from each condition

mild_vs_ctrl <- read.csv("/path/to/DEresults/progenitor_mild.csv")
mild_sig <- mild_vs_ctrl[mild_vs_ctrl$p_val_adj<0.05 & abs(mild_vs_ctrl$avg_log2FC)>0.5 ,]
mild_sig <- mild_sig[mild_sig$pct.1 > 0.25 | mild_sig$pct.2 > 0.25, ]
mild_top10 <- mild_sig$X[order(abs(mild_sig$avg_log2FC),decreasing = T)][c(1:10)]
#write.table(mild_sig, "/path/to/DEresults/progenitor_mild_significant.csv",row.names = F, quote = F,sep=",")

moderate_vs_ctrl <- read.csv("/path/to/DEresults/progenitor_moderate.csv")
moderate_sig <- moderate_vs_ctrl[moderate_vs_ctrl$p_val_adj<0.05 & abs(moderate_vs_ctrl$avg_log2FC)>0.5 ,]
moderate_sig <- moderate_sig[moderate_sig$pct.1 > 0.25 | moderate_sig$pct.2 > 0.25, ]
#write.table(moderate_sig, "/path/to/DEresults/progenitor_moderate_significant.csv",row.names = F, quote = F,sep=",")
moderate_top10 <- moderate_sig$X[order(abs(moderate_sig$avg_log2FC),decreasing = T)][c(1:10)]

severe_vs_ctrl <- read.csv("/path/to/DEresults/progenitor_severe.csv")
severe_sig <- severe_vs_ctrl[severe_vs_ctrl$p_val_adj<0.05 & abs(severe_vs_ctrl$avg_log2FC)>0.5 ,]
severe_sig <- severe_sig[severe_sig$pct.1 > 0.25 | severe_sig$pct.2 > 0.25, ]
severe_top10 <- severe_sig$X[order(abs(severe_sig$avg_log2FC),decreasing = T)][c(1:10)]
#write.table(severe_sig, "/path/to/DEresults/progenitor_severe_significant.csv",row.names = F, quote = F,sep=",")

# Get list of DE genes that need to be included in the heatmap
DE_genes <- c(mild_sig$X,moderate_sig$X,severe_sig$X)
DE_genes <- DE_genes[!duplicated(DE_genes)]

# Obtain log2FCs for each cluster for the selection of DE genes
mild <- mild_vs_ctrl[mild_vs_ctrl$X %in% DE_genes,c("X","avg_log2FC")]
colnames(mild)[2] <- "mild"

moderate <- moderate_vs_ctrl[moderate_vs_ctrl$X %in% DE_genes,c("X","avg_log2FC")]
colnames(moderate)[2] <- "moderate"

severe <- severe_vs_ctrl[severe_vs_ctrl$X %in% DE_genes,c("X","avg_log2FC")]
colnames(severe)[2] <- "severe"



# merge dfs
library(tidyverse)
dfs<-list(mild, moderate, severe) %>% reduce(inner_join,by="X")
rownames(dfs)<-dfs$X

# Create log2FC heatmap
hm_DEG <- as.matrix(dfs[,-1])

#reorder the rows to match with annotation labels
library(ComplexHeatmap)
rows_DEG <- row_order(Heatmap(hm_DEG,name="log2FC",column_names_gp  = gpar(fontsize=8),column_names_rot = 45))
hm_DEG <- hm_DEG[rows_DEG,]

# subset genes for labeling in the heatmap
lab <- rownames(hm_DEG)
lab_genes <- c(mild_top10,moderate_top10,severe_top10)
 #remove mitochondrial genes, pseudogenes and XIST

lab_genes <- lab_genes[-grep("RPS|RPL|XIST|MT-|AC|AL",lab_genes)]
lab[!(lab %in% lab_genes)] <- ""
lab_genes <- unique(lab_genes)

# Plot heatmap
library(colorRamp2)
col_fun_rna = colorRamp2(c(0,0.25, 0.5,0.75, 1), c("#1B4D68","#28739C","#E4E4E4","#F7AB64","#C7361B"))
col_fun_prot = colorRamp2(c(0,0.25, 0.5,0.75, 1), c("#003E65","#3A98CC","#E4E4E4","#F7AB64","#C7361B"))

#pdf("/path/to/figures/Fig2E_DEG_hm.pdf",width=3,height=3)
Heatmap(hm_DEG,name="log2FC", col=col_fun_rna(seq(0,1,length=100)),show_row_dend=F,cluster_columns = FALSE,column_names_gp  = gpar(fontsize=8),column_names_rot = 45,heatmap_legend_param = list(title_gp = gpar(fontsize = 7),labels_gp=gpar(fontsize=6))) + rowAnnotation(link = anno_mark(at = which(rownames(hm_DEG)==lab), 
        labels = lab[lab !=""], padding = unit(1, "mm"),link_gp = gpar(lwd = 0.1) ,labels_gp = gpar(fontsize = 7)))
#dev.off()

#venn.diagram(x=list(mild_sig$X,moderate_sig$X,severe_sig$X), category.names = c("mild", "moderate", "severe"),  fill = c("#F7AB64","#C7361B","#880000"),fontfamily ="calibri",cat.fontfamily = "calibri", cex = 2, cat.cex = 1.3, "/path/to/figures/venn.png")

```


```{r Venn Diagram, warning=FALSE, results='hide'}
mod_vs_mild <- read.csv("/path/to/DEresults/progenitor_moderate_vs_mild.csv")
mod_vs_mild_sig <- mod_vs_mild[mod_vs_mild$p_val_adj<0.05 & abs(mod_vs_mild$avg_log2FC)>0.5 ,]
mod_vs_mild_sig <- mod_vs_mild_sig[mod_vs_mild_sig$pct.1 > 0.25 | mod_vs_mild_sig$pct.2 > 0.25, ]
write.table(mod_vs_mild_sig, "/path/to/DEresults/progenitor_moderate_vs_ mild_significant.csv",row.names = F, quote = F,sep=",")

sev_vs_mild <- read.csv("/path/to/DEresults/progenitor_severe_vs_mild.csv")
sev_vs_mild_sig <- sev_vs_mild[sev_vs_mild$p_val_adj<0.05 & abs(sev_vs_mild$avg_log2FC)>0.5 ,]
sev_vs_mild_sig <- sev_vs_mild_sig[sev_vs_mild_sig$pct.1 > 0.25 | sev_vs_mild_sig$pct.2 > 0.25, ]
write.table(sev_vs_mild_sig, "/path/to/DEresults/progenitor_severe_vs_ mild_significant.csv",row.names = F, quote = F,sep=",")

sev_vs_mod <- read.csv("/path/to/DEresults/progenitor_severe_vs_moderate.csv")
sev_vs_mod_sig <- sev_vs_mod[sev_vs_mod$p_val_adj<0.05 & abs(sev_vs_mod$avg_log2FC)>0.5 ,]
sev_vs_mod_sig <- sev_vs_mod_sig[sev_vs_mod_sig$pct.1 > 0.25 | sev_vs_mod_sig$pct.2 > 0.25, ]
write.table(sev_vs_mod_sig, "/path/to/DEresults/progenitor_severe_vs_ moderate_significant.csv",row.names = F, quote = F,sep=",")


x <- list(
  mild_ctrl = mild_sig$X, 
  moderate_mild = mod_vs_mild_sig$X,
  moderate_ctrl = moderate_sig$X, 
  severe_ctrl = severe_sig$X,
  severe_mild = sev_vs_mild_sig$X,
  severe_moderate = sev_vs_mod_sig$X
  )

ihml <- Reduce(union,  x) 

test <- cbind(ihml, ihml%in%mild_sig$X, ihml%in%mod_vs_mild_sig$X, ihml%in%moderate_sig$X, 
              ihml%in%severe_sig$X, ihml%in%sev_vs_mild_sig$X,ihml%in%sev_vs_mod_sig$X)

rownames(test) <- test[,1]
test <- test[,-1]
colnames(test) <- c("mild_vs_ctrl","moderate_vs_mild","moderate_vs_ctrl","severe_vs_ctrl","severe_vs_mild","severe_vs_moderate")
test[test==TRUE]<-1
test[test==FALSE]<-0

test <- data.frame(test)
test[,1] <- as.numeric(test[,1])
test[,2] <- as.numeric(test[,2])
test[,3] <- as.numeric(test[,3])
test[,4] <- as.numeric(test[,4])
test[,5] <- as.numeric(test[,5])
test[,6] <- as.numeric(test[,6])

testmat <- as.matrix(test)

library(RColorBrewer)
#coul <- colorRampPalette(c("#C7361B","#D0D0CF"))(2)
coul <- colorRampPalette(c("#C7361B","#FCD8C1"))(2)

png("/path/to/figures/all_genes.png", width = 4.50,height=6.9,units="in",res=1200)
heatmap(testmat, scale="none",col=coul, Colv = NA, Rowv = NA)
dev.off()

colnames(df) <- c("mild_vs_ctrl","moderate_vs_mild","moderate_vs_ctrl","severe_ctrl","severe_mild","severe_moderate")
write.table(df, "/path/to/DEresults/all_gene_list.csv",row.names = F, quote = F,sep=",")


```

- Figure 3D

## Violin plots GO terms 

```{r define function, warning=FALSE, results='hide'}
library(AnnotationDbi)
add_GO_perc <- function(data, GO_term){
  go_id = GOID(GOTERM[Term(GOTERM) == GO_term])
  allegs <- get(go_id, org.Hs.egGO2ALLEGS)
  genes <- unlist(mget(allegs, org.Hs.egSYMBOL))
  genes <- unique(genes)        
  genes <- genes[genes %in% rownames(data)]
  PercentageFeatureSet(object=data, features=genes, col.name = GO_term)
}  
```

```{r subset sc object, warning=FALSE, results='hide'}
DefaultAssay(sc) <- "RNA"
sc@meta.data$cluster_condition <- paste0(sc@meta.data$condition,sc@meta.data$seurat_clusters)
Idents(sc) <- "cluster_condition"

RG_tRG <- FindMarkers(sc,ident.1 = "Severe3", ident.2 = "Ctrl0", min.pct = 0.25,assay = "RNA", log2threshold=0.5)
RG_tRG <- RG_tRG[RG_tRG$p_val_adj<.05&abs(RG_tRG$avg_log2FC>0.5),]
head(RG_tRG,100)

RG_tRG$gene <- rownames(RG_tRG)

#write.table(RG_tRG, "/path/to/DEresults/RG_t-RG_marker.txt", row.names = F,sep=";",quote=F)

library(org.Hs.eg.db)
hs <- org.Hs.eg.db
ann_RGtRG <- AnnotationDbi::select(hs, 
       keys = RG_tRG$gene,
       columns = c("ENTREZID", "SYMBOL"),
       keytype = "SYMBOL")
library(clusterProfiler)
go.obj <- enrichGO(ann_RGtRG$ENTREZID,OrgDb = "org.Hs.eg.db",ont="BP")
go.res <- go.obj@result

#write.table(go.res,"/path/to/DEresults/RG_t-RG_GO.txt", row.names = F,sep=";",quote=F)

# Comparison only in Severe vs. Ctrl, cluster 3 = t-RG, cluster 0 = RG
sc2 <- subset(sc, idents=c("Severe3","Ctrl0"))
Idents(sc2) <- "cell_types"
```

## RG and t-RG clusters - neurogenesis-related terms - Figure 3D
```{r GO plots, warning=FALSE, results='hide'}
GO_terms <- c("neuron differentiation","neuron fate commitment","regulation of neuron differentiation","epithelial tube formation")

for(i in seq_along(GO_terms)){
  sc2 <- add_GO_perc(sc2, GO_terms[i])
}

titles <- c("neuron differentiation","neuron fate commitment","regulation of neuron differentiation","epithelial tube formation")
GO_terms_no_space <- str_replace_all(GO_terms, " ", ".")
plots <- map2(GO_terms_no_space, titles, ~VlnPlot(sc2, features=.x, split.by = "cell_types", pt.size=0.1)+ ggtitle(.y))

wil_t <- list()
for(i in GO_terms_no_space){
  df <- data.frame(GO=sc2[[i]], clusters=sc2$cell_types)
  wil_t[[i]] <- wilcox.test(x=filter(df, clusters=="RG")[,1], y=filter(df, clusters=="t-RG")[,1])
}

# p-values 
wil_t$neuron.differentiation$p.value #2.520929e-15
wil_t$generation.of.neurons$p.value #3.649961e-14
wil_t$neuron.fate.commitment$p.value #3.625523e-61
wil_t$regulation.of.neuron.differentiation$p.value #1.506357e-190

plots[[1]] <- plots[[1]]+
  annotate("text", x=1.5, y=max(sc2[[GO_terms_no_space[1]]])+max(sc2[[GO_terms_no_space[1]]])*0.1, label="***", size=10)+
  ylim(min(sc2[[GO_terms_no_space[1]]])-min(sc2[[GO_terms_no_space[1]]])*0.15,max(sc2[[GO_terms_no_space[1]]])+max(sc2[[GO_terms_no_space[1]]])*0.15)
plots[[2]] <- plots[[2]]+
  annotate("text", x=1.5, y=max(sc2[[GO_terms_no_space[2]]])+max(sc2[[GO_terms_no_space[2]]])*0.1, label="***", size=10)+
  ylim(min(sc2[[GO_terms_no_space[2]]])-min(sc2[[GO_terms_no_space[2]]])*0.15,max(sc2[[GO_terms_no_space[2]]])+max(sc2[[GO_terms_no_space[2]]])*0.15)
plots[[3]] <- plots[[3]]+
  annotate("text", x=1.5, y=max(sc2[[GO_terms_no_space[3]]])+max(sc2[[GO_terms_no_space[3]]])*0.1, label="***", size=10)+
  ylim(min(sc2[[GO_terms_no_space[3]]])-min(sc2[[GO_terms_no_space[3]]])*0.15,max(sc2[[GO_terms_no_space[3]]])+max(sc2[[GO_terms_no_space[3]]])*0.15)
plots[[4]] <- plots[[4]]+
  annotate("text", x=1.5, y=max(sc2[[GO_terms_no_space[4]]])+max(sc2[[GO_terms_no_space[4]]])*0.1, label="***", size=10)+
  ylim(min(sc2[[GO_terms_no_space[4]]])-min(sc2[[GO_terms_no_space[4]]])*0.15,max(sc2[[GO_terms_no_space[4]]])+max(sc2[[GO_terms_no_space[4]]])*0.15)


vln <- wrap_plots(plots, guides="collect", nrow=2)& 
  theme(axis.title.x=element_blank(), axis.text.y=element_text(size=10), plot.title = element_text(size=8.5), legend.text = element_text(size=10), axis.title.y=element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())&
  ylab("Proportion of feature set")&
  scale_fill_manual(values=c("#CC5500","#880000"), labels=c("RG","t-RG"))

ggsave("/path/to/figures/VlnPlot.png", vln, width= 5, height = 5, dpi=600)
```


```{r GO plots apoptosis, warning=FALSE, results='hide'}

GO_terms_2 <- c("positive regulation of apoptotic process","glial cell apoptotic process","cell death")

for(i in seq_along(GO_terms_2)){
  sc <- add_GO_perc(sc, GO_terms_2[i])
}


titles <- c("positive regulation of apoptotic process","glial cell apoptotic process","cell death")
GO_terms_no_space2 <- str_replace_all(GO_terms_2, " ", ".")

FeaturePlot(sc, features = c("positive.regulation.of.apoptotic.process","glial.cell.apoptotic.process","cell.death"))

replicates <- c("C1","C7","P1","P2","P3","P4","P5","P6")

perc_tab <- data.frame(replicate = replicates,
                       condition = c("control","control","mild","mild","moderate","moderate","severe","severe"),
                       mean_apop = rep(0, times = 8),
                       sd_apop = rep(0, times = 8),
                       mean_glia = rep(0, times = 8),
                       sd_glia = rep(0, times = 8),
                       mean_celld = rep(0, times = 8),
                       sd_celld = rep(0, times = 8))

for (i in replicates) {
  perc_tab[replicates==i,"mean_apop"] <- mean(sc@meta.data$positive.regulation.of.apoptotic.process[sc@meta.data$replicate==i],na.rm=T)
  perc_tab[replicates==i,"sd_apop"] <-  sd(sc@meta.data$positive.regulation.of.apoptotic.process[sc@meta.data$replicate==i])
  
  perc_tab[replicates==i,"mean_glia"] <- mean(sc@meta.data$glial.cell.apoptotic.process[sc@meta.data$replicate==i])
  perc_tab[replicates==i,"sd_glia"] <-  sd(sc@meta.data$glial.cell.apoptotic.process[sc@meta.data$replicate==i])
  
  perc_tab[replicates==i,"mean_celld"] <- mean(sc@meta.data$cell.death[sc@meta.data$replicate==i])
  perc_tab[replicates==i,"sd_celld"] <-  sd(sc@meta.data$cell.death[sc@meta.data$replicate==i])
}

write.table(perc_tab,"/path/to/apo_table.csv", row.names = F, sep=",")


apo1 <- FeaturePlot(sc, features = GO_terms_no_space2, ncol=1)
ggsave("/path/to/figures/FigRev1_ApoPlot.png", apo1, width= 6.5, height = 15, dpi=600)

#p_apop<-ggplot(perc_tab, aes(x=replicate, y=mean_apop, fill=condition)) +  
#  geom_point()+ ggtitle("positive regulation of apoptotic process") +
#  geom_errorbar(aes(ymin=mean_apop-sd_apop, ymax=mean_apop+sd_apop), width=.2, 
#                position=position_dodge(0.05)) + theme_classic()

p_apop<-ggplot(perc_tab, aes(x=replicate, y=mean_apop, fill=condition)) + 
  geom_bar(position=position_dodge(), stat="identity", fill=c("gray93","gray93","gray68","gray68","gray43","gray43","gray23","gray23"),
           colour='black') + ggtitle("positive regulation of apoptotic process") +
  geom_errorbar(aes(ymin=mean_apop-sd_apop, ymax=mean_apop+sd_apop), width=.2) + theme_classic() +
  theme(text = element_text(size = 20))+ylab("GO score apoptotic process") + xlab("")

#p_celld<-ggplot(perc_tab, aes(x=replicate, y=mean_celld)) +  
#  geom_point()+ ggtitle("cell death") +
#  geom_errorbar(aes(ymin=mean_celld-sd_celld, ymax=mean_celld+sd_celld), width=.2, 
#                position=position_dodge(0.05)) + theme_classic()

p_celld<-ggplot(perc_tab, aes(x=replicate, y=mean_celld, fill=condition)) + 
  geom_bar(position=position_dodge(), stat="identity", fill=c("gray93","gray93","gray68","gray68","gray43","gray43","gray23","gray23"),
           colour='black') + ggtitle("cell death") +
  geom_errorbar(aes(ymin=mean_celld-sd_celld, ymax=mean_celld+sd_celld), width=.2) + theme_classic()+
  theme(text = element_text(size = 20))+ylab("GO score cell death") + xlab("")

ggsave("/path/to/figures/ApoPlot2.png", ggarrange(p_apop, p_celld, ncol=1, common.legend = T, legend = "bottom"), width= 9, height = 8, dpi=600)


m_apop <- lm(positive.regulation.of.apoptotic.process ~ condition, data = sc@meta.data)
parameters::model_parameters(anova(m_apop)) #Cohens d = 0.08

m_glia <- lm(glial.cell.apoptotic.process ~ condition, data = sc@meta.data)
parameters::model_parameters(anova(m_glia)) #Cohens d = 0.037

m_celld <- lm(cell.death ~ condition, data = sc@meta.data)
parameters::model_parameters(anova(m_celld)) #Cohens d = 0.114
```


```{r subset sc object for GO plots, warning=FALSE, results='hide'}
DefaultAssay(sc) <- "RNA"
Idents(sc) <- "DE_progenitor"

sc3 <- subset(sc, idents=c("Ctrl_progenitor","Mild_progenitor","Moderate_progenitor","Severe_progenitor"))
Idents(sc3) <- "condition"
```


- Figure 3G

# WNT gene heatmap in CyP cluster 

```{r define stars function, warning=FALSE}
stars_gen <- function(x){
  stars <- c( "***", "**", "*","")
  vec <- c(0, 0.001, 0.01, 0.05,1.01)
  i <- findInterval(x, vec)
  stars[i]
}
```

# Generate DE gene lists for CyP cluster
```{r DE 1, results='hide', warning=FALSE, results='hide'}
Idents(sc)<- "cluster_condition"

# Cluster 1 are CyP cells
markers_Mild_vs_Ctrl_1 <- FindMarkers(sc,ident.1 = "Mild1" , ident.2 = "Ctrl1", min.pct = 0.25,assay = "RNA", logfc.threshold = 0)
markers_Moderate_vs_Ctrl_1 <- FindMarkers(sc,ident.1 = "Moderate1" , ident.2 = "Ctrl1", min.pct = 0.25,assay = "RNA", logfc.threshold = 0)
markers_Severe_vs_Ctrl_1 <- FindMarkers(sc,ident.1 = "Severe1" , ident.2 = "Ctrl1", min.pct = 0.25,assay = "RNA", logfc.threshold = 0)


```

# Specify marker genes
```{r gene lists, results='hide', warning=FALSE}
list <- c("AXIN1","BCL9","BCL9L","CBY1","CSNK1E","CTDNEP1","CTNNB1","DACT3","DISC1","DIXDC1","DKK1","DVL1","DVL1P1","DVL2",    "DVL3","EDA","EDN1","EDNRA","EDNRB","EXT1","FGF2","FGF9","FOXD1","FOXO1","FOXO3","FRAT1" ,"FRZB","FZD1","FZD10","FZD2","FZD3" ,   "FZD4","FZD5","FZD6","FZD7","FZD8","FZD9","GPC3","HESX1","ISL1","JUP","KLF4","LATS2","LEF1","LRP5","LRP6","LRRK2","NR4A2","PLt-RG3",   "PTEN" ,"PYGO1","PYGO2","RARG","RECK","RSPO2","RSPO3","RYK","SDC1","SFRP1","SFRP2","SFRP4","SFRP5","SHH","SIAH1","SIAH2","SOST",    "SOSTDC1","SOX9","STK3" , "STK4","TCF7","TCF7L1","TCF7L2","WNT1" ,"WNT10A","WNT10B","WNT11" ,"WNT16" ,"WNT2","WNT2B","WNT3",    "WNT3A" ,"WNT4","WNT5A","WNT5B","WNT6","WNT7A","WNT7B","WNT8A","WNT8B","WNT9A","WNT9B","YAP1")
```

# Generate heatmap

```{r heatmapsWntcanonicalCyP, results='hide', warning=FALSE}

list_Mild_vs_Ctrl_1 <- markers_Mild_vs_Ctrl_1[rownames(markers_Mild_vs_Ctrl_1) %in% list,2:5]
colnames(list_Mild_vs_Ctrl_1) <- paste0("Mild_",colnames(list_Mild_vs_Ctrl_1))
list_Moderate_vs_Ctrl_1 <- markers_Moderate_vs_Ctrl_1[rownames(markers_Moderate_vs_Ctrl_1) %in% list,2:5]
colnames(list_Moderate_vs_Ctrl_1) <- paste0("Moderate_",colnames(list_Moderate_vs_Ctrl_1))
list_Severe_vs_Ctrl_1 <- markers_Severe_vs_Ctrl_1[rownames(markers_Severe_vs_Ctrl_1) %in% list,2:5]
colnames(list_Severe_vs_Ctrl_1) <- paste0("Severe_",colnames(list_Severe_vs_Ctrl_1))

# Subset shared genes CyP
shared <- Reduce(intersect, list(rownames(list_Mild_vs_Ctrl_1),rownames(list_Moderate_vs_Ctrl_1),rownames(list_Severe_vs_Ctrl_1)))


list_Mild_vs_Ctrl_1 <- list_Mild_vs_Ctrl_1[rownames(list_Mild_vs_Ctrl_1) %in% shared,]
list_Moderate_vs_Ctrl_1 <- list_Moderate_vs_Ctrl_1[rownames(list_Moderate_vs_Ctrl_1) %in% shared,]
list_Severe_vs_Ctrl_1 <- list_Severe_vs_Ctrl_1[rownames(list_Severe_vs_Ctrl_1) %in% shared,]
cluster1 <- cbind(list_Mild_vs_Ctrl_1,list_Moderate_vs_Ctrl_1,list_Severe_vs_Ctrl_1)


# Select log2FC values
cluster1_log2FC <- as.matrix(cluster1[,c(1,5,9)])
colnames(cluster1_log2FC) <- c("Mild","Moderate","Severe")


# Generate information on significance
cluster1_p <- cluster1[,c(4,8,12)]
colnames(cluster1_p) <- c("Mild","Moderate","Severe")
for(i in colnames(cluster1_p)){
cluster1_p[,i] <- stars_gen(cluster1_p[,i])}
cluster1_p <- as.matrix(cluster1_p)


cluster1_log2FC <- cluster1_log2FC[order(rownames(cluster1_log2FC)),]

cluster1_p <- cluster1_p[order(rownames(cluster1_p)),]

# Plot heatmap
png("/path/to/figures/Wnt_CyP.png", width = 300,height=700)
pheatmap(cluster1_log2FC, breaks=seq(-1.5, 1.5, length.out=100),color=col_fun_rna(seq(0,1,length=100)),cluster_rows = F,cluster_cols = F,fontsize = 16,display_numbers = cluster1_p, cellwidth = 20,cellheight = 20)
dev.off()
```
